Cubical Agda features

### 2021-09-09 slides.lagda.tex
fixed grammar and finished slides

%Information to be included in the title page:
\title{Using quotient types in Agda}
\author{Guilherme Silva}
\institute{Scottish Programming Languages and Verification Summer School}
\date{17th August 2021}

@@

\begin{frame}
  \frametitle{Motivation}
  In simple type theories with quotient types,
  it is possible to create two distinct elements and after
  make them equal.

@@

@@

  open import Cubical.Data.Unit
  open import Cubical.Data.Bool
  open import Cubical.Data.Int
  open import Cubical.Data.Nat hiding (_·_)

@@

\begin{frame}
  \frametitle{Equalities}
  In cubical type theory, equality is a function between from an interval to a set.
  \begin{code}
  tt≡tt : tt ≡ tt
  tt≡tt i = tt

  refl' : (x : A) → x ≡ x
  refl' x i = x
  \end{code}

\begin{frame}
  \frametitle{Equalities}
  It is also possible to invert the path of equality.
  If the interval i goes from i0 to i1, the interval \~{} i goes from i1 to i0.

  \begin{code}
  sym' : x ≡ y → y ≡ x
  sym' p i = p (~ i)

  ext : (f g : A → B) → (∀ x → f x ≡ g x) → f ≡ g
  ext f g p i x = -- wants that when i = i0, f x
                  --        and when i = i1, g x
                  p x -- is the same of (f x ≡ g x)
                  i
  \end{code}

@@

  In truncation type, every element of it is equal.
  Therefore we can define Bool≡ as ∥ Bool ∥.

@@

\begin{frame}
  \frametitle{Making equalities equal}
  To make the equality equal, it is necessary when declaring a data type to state that it is a set.
  \begin{code}
  isProp : Type ℓ → Type ℓ
  isProp A = (x y : A) → x ≡ y

@@

\begin{frame}
  \frametitle{Rational Numbers}
  Rational numbers are defined by numerator and a denominator different from zero.
  Two rational numbers $ \frac{p_1}{q_1} $ and $ \frac{p_2}{q_2} $ are equal when $ p_1 \times q_2 \equiv p_2 \times q_1 $
  \begin{code}
  data ℚ : Type where
    _/_[_] : (p : ℤ) (q : ℤ) → ¬ (q ≡ pos 0) → ℚ
    path : ∀ p₁ q₁ p₂ q₂ {pr₁ pr₂} → (p₁ · q₂) ≡ (p₂ · q₁)
      → p₁ / q₁ [ pr₁ ] ≡ p₂ / q₂ [ pr₂ ]
    trunc : isSet ℚ

@@

\begin{frame}
  \frametitle{Integers}
  Another way to define integers is making the positive and negative zero equals.
  \begin{code}
  data ℤ' : Type where
    pos' : ℕ → ℤ'
    neg' : ℕ → ℤ'
    0+-≡ : pos' 0 ≡ neg' 0
  \end{code}
\end{frame}

\begin{frame}
  \frametitle{Sets}
  One way of defining sets is to create a list where you can swap elements and remove them when they are equal.
  \begin{code}
  infixr 20 _∷_
  data LFSet (A : Type ℓ) : Type ℓ where
    []    : LFSet A
    _∷_   : (x : A) → (xs : LFSet A) → LFSet A
    dup   : ∀ x xs   → x ∷ x ∷ xs ≡ x ∷ xs
    comm  : ∀ x y xs → x ∷ y ∷ xs ≡ y ∷ x ∷ xs
    trunc : isSet (LFSet A)

@@

\begin{frame}
  \vspace*{36 pt}
  \begin{center}
  {\Huge Questions?}
  \end{center}
\end{frame}

>>>>

%Information to be included in the title page:
\title{Cubical Agda features}
\author{Guilherme Silva}
\date{11th September 2021}

@@

\section{Cubical Agda}

\begin{frame}
  \frametitle{New Features}
  The most important features in Cubical Agda are:

  \begin{itemize}
    \item New Equality Type
    \item Quotient types
  \end{itemize}
\end{frame}

\section{Equalities}

\begin{frame}
  \frametitle{Equalities}
  There are multiple definitions of equalities and the most important are:
  \begin{itemize}
    \item Martin-Löf
    \item Leibniz
    \item Cubical
  \end{itemize}
\end{frame}

  module slides where

  open import Agda.Primitive.Cubical

  open import Cubical.Data.Unit
  open import Cubical.Data.Bool
  open import Cubical.Data.Int
  open import Cubical.Data.Nat hiding (_·_)
  open import Cubical.Data.Prod
  open import Cubical.Foundations.Isomorphism
  \end{code}

\begin{frame}
  \frametitle{Martin-Löf equality}
  At the begin of Agda and in most theorems proves, equality is given by Martin-Löf's definition:
  \begin{code}
  module Martin-Löf {ℓ} {A : Set ℓ} where

    data _≡_ (x : A) : A → Set ℓ where
      refl : x ≡ x
  \end{code}
\end{frame}

\begin{frame}
  \frametitle{Martin-Löf pattern match}
  This equality is very convenient in proof assistances like Agda because it is possible to pattern match using them:

  \begin{code}
    private variable
      x y z : A

    sym  : x ≡ y → y ≡ x
    sym refl = refl

    trans : x ≡ y → y ≡ z → x ≡ z
    trans refl refl = refl
  \end{code}
\end{frame}

\begin{frame}
  \frametitle{Martin-Löf problem}

  But the problem of this equality is that it does not handle extensionality and other axioms very well:

  \begin{code}
  module FunExt {ℓ ℓ'} {A : Set ℓ} {B : Set ℓ'} where
    open Martin-Löf

    funExt-Type = {f g : A → B}
      → ((x : A) → f x ≡ g x) → f ≡ g
  \end{code}
\end{frame}

\begin{frame}
  \frametitle{Cubical Equality}
  To solve this problem, Agda adopted cubical type theory that equality is a function from the path to type:

  \begin{code}
  module CubicalEquality {ℓ} {A : Set ℓ} where
    postulate
      PathP : (A : I → Set ℓ) → A i0 → A i1 → Set ℓ

    _≡_ : A → A → Set ℓ
    _≡_ = PathP λ _ → A
  \end{code}
\end{frame}

\begin{frame}
  \frametitle{Cubical Equality Operators}
  From this equality, I will define reflection, symmetry and extensionality:

  \begin{code}
  module CubicalResults {ℓ ℓ'} {A : Set ℓ} {B : Set ℓ'} where
    open import Cubical.Core.Primitives

    private variable
      x y z : A

    refl : x ≡ x
    refl {x = x} = λ _ → x

    sym : x ≡ y → y ≡ x
    sym p i = p (~ i)

    funExt : {f g : A → B}
      → ((x : A) → f x ≡ g x) → f ≡ g
    funExt p i x = p x i
  \end{code}

  The operator $\sim$ invert the interval. If the interval \AgdaBound{i} goes from
  \AgdaBound{i0} to \AgdaBound{i1}, the interval $\sim i$ goes from \AgdaBound{i1} to \AgdaBound{i0}.
\end{frame}

\begin{frame}
  \frametitle{Cubical equality computational behavior}
  Another advantage of cubical equality is that it has a computational behavior.
  If there is a term that has a type that two types are equal, in this term, there is information of how they are equal.
  So this term represents the isomorphism of these two types.
\end{frame}

\begin{frame}
  \frametitle{Bool example}

  In this example, I will show the equality of two Bools:

  \begin{code}
  module CubicalIsomorphism where
    open import Cubical.Foundations.Prelude

    IsoBool : Iso Bool Bool
    IsoBool = iso not not ¬¬b ¬¬b
      where
        ¬¬b : ∀ b → not (not b) ≡ b
        ¬¬b false = refl
        ¬¬b true = refl

    Bool≡Bool : Bool ≡ Bool
    Bool≡Bool = isoToPath IsoBool

    _ : (transport Bool≡Bool false ≡ true)
      × (transport Bool≡Bool true ≡ false)
    _ = refl , refl

  \end{code}
\end{frame}

\section{Quotient Types}

\begin{frame}
  \frametitle{Quotient Types' Motivation}
  In simple type theories with quotient types,
  it is possible to create two distinct elements and after
  make them equal.

@@

\begin{frame}
  \frametitle{Imports}
  \begin{code}

@@

@@

@@

  In truncation type, every element of it is equal.
  Therefore we can define Bool≡ as ∥ Bool ∥:

@@

\begin{frame}
  \frametitle{Making equalities equal}
  To make the equality equal, it is necessary when declaring a data type to state that it is a set:
  \begin{code}
  isProp : Type ℓ → Type ℓ
  isProp A = (x y : A) → x ≡ y

@@

\begin{frame}
  \frametitle{Rational Numbers}
  Rational numbers are defined by numerator and a denominator different from zero.
  Two rational numbers $ \frac{p_1}{q_1} $ and $ \frac{p_2}{q_2} $ are equal when $ p_1 \times q_2 \equiv p_2 \times q_1 $:
  \begin{code}
  data ℚ : Type where
    _/_[_] : (p : ℤ) (q : ℤ) → ¬ (q ≡ pos 0) → ℚ
    path : ∀ p₁ q₁ p₂ q₂ {pr₁ pr₂} → (p₁ · q₂) ≡ (p₂ · q₁)
      → p₁ / q₁ [ pr₁ ] ≡ p₂ / q₂ [ pr₂ ]
    trunc : isSet ℚ

@@

\begin{frame}
  \frametitle{Integers}
  Another way to define integers is making the positive and negative zero equals:
  \begin{code}
  data ℤ' : Type where
    pos' : ℕ → ℤ'
    neg' : ℕ → ℤ'
    0+-≡ : pos' 0 ≡ neg' 0
  \end{code}
\end{frame}

\begin{frame}
  \frametitle{Sets}
  One way of defining sets is to create a list where you can swap elements and remove them when they are equal:
  \begin{code}
  infixr 20 _∷_
  data LFSet (A : Type ℓ) : Type ℓ where
    []    : LFSet A
    _∷_   : (x : A) → (xs : LFSet A) → LFSet A
    dup   : ∀ x xs   → x ∷ x ∷ xs ≡ x ∷ xs
    comm  : ∀ x y xs → x ∷ y ∷ xs ≡ y ∷ x ∷ xs
    trunc : isSet (LFSet A)

@@
